---
title: "G Theory for fMRI Batch Investigations"
author: "Eric Bridgeford"
date: "October 12, 2017"
output: html_document
---


# Setting

## Task

Given:

+ $n$ samples of data-tuples, $\left\{(g_i, y_i, z_i)\right\}_{i=1}^n$.
+ A graph, $g_i \in G$, where $g_i  = (E, V, w)$ for $N=|V|$ regions of interest and $w(v_i, v_j) = w_{ij}$.
+ A set of fixed-effects, $y_i \in Y$, where $Y$ is the set of variables we expect will have an impact on connectome-to-connectome variation comprised of signal.
+ A set of random-effects, $z_i \in Z$, where $Z$ is the set of variables we expect will have an impact on connectome-to-connectome variation comprised of batch-effects.

Characterize the composition of signal to batch-effect present in the given connectomes, assuming both the fixed-effects and random-effects can be characterized as random inputs in a standard mixed-effects model.

## Statistical Model

Assume that the desired signal at a given edge can be modelled as a linear combination of the random mixed-effects under the Generalizability theory (G-theory) model:

\begin{align*}
  w_{ij} = Zu + \epsilon
\end{align*}

+ $w_{ij}$ as above
+ $Z$ is the known design matrix relating the observations $y$ to $u$
+ $u$ is the unknown vector of random-effects with $\mathbb{E}\left[u\right] = 0$ and $var(u) = G$
+ $\epsilon$ is the unknown vector of random errors, where $\mathbb{E}\left[\epsilon\right] = 0$ and $var(\epsilon) = R$.

## Statistical Goal

Given that the joint density of $y$ and $u$ can be written $f(y, u) = f(y | u) f(u)$, assuming the normality and independence of our random-effects and errors:

\begin{align*}
  \begin{bmatrix}
    u \\ \epsilon
  \end{bmatrix} = \mathcal{N}\left(0, diag\left(G, R\right)\right)
\end{align*}

where all covariance terms are $0$, we can solve this system using linear regression and estimate variance terms from unobserved nuisance parameters in the joint likelihood. 

Our goal is to characterize the variance components $\sigma_i$ from mixed-effects combination $i$. Under our given model, we are chiefly interested in several quantities:

### Confidence that a given mixed-effect is significant

We are interested in the quantity of whether a given mixed-effect combinations is significant. To accomplish this, we can formulate the hypothesis of whether $H_0 := \mu_1 = ... = \mu_K$; that is, the mean activity at each edge $\mu_i$ are the same across the $i$ groupings of our given mixed-effect of interest, with $H_A := \textrm{the means are not equal}$. This can be examined using the F-test statistic:

\begin{align*}
  F &= \frac{\textrm{between-group variability}}{\textrm{within-group varaibility}} \\
  &= \frac{\sum_{i=1}^K n_i (\bar{Y}_i - \bar{Y})^2/(K - 1)}{\sum_{i=1}^K\sum_{j=1}^{n_i} (Y_{ij} - \bar{Y_{i})^2}/(N - K)}
\end{align*}

We can the compute the $p-$value where:
\begin{align*}
  p = \mathbb{P}\left\{\textrm{Falsely reject $H_0$ for $H_A \; | \; H_0$ is true}\right\}
\end{align*}

### Proportion of Total Variability accounted by a given mixed-effect

We are interested in the proportion of the variance accounted for by each variance component:

\begin{align*}
  V &= \frac{\sigma_i^2}{\sum_{j \in M}\sigma_j^2}
\end{align*}

where $M$ is the set of possible facets and between-facet interactions.

### D-Coefficient

For the inivariate case, we can estimate the dependability $D$-coefficients:

\begin{align*}
  \phi_{UV}(X) = \frac{\sigma^2_p(x)}{\sigma_p(x)^2 + \sum_{i\in M \ p} \frac{\sigma_i^2}{n_i}}
\end{align*}

where $n_i$ are the number of conditions facet $i$ can occupy, and $p$ are our test-subjects in our investigation.

### Multivariate Test-Retest Reliability


\begin{align*}
  \phi_{MV}(X) = \frac{\sum_{x=1}^X\sigma^2_p(x)}{\sum_{x=1}^X\left(\sigma_p(x)^2 + \sum_{i\in M \ p} \frac{\sigma_i^2}{n_i}\right)}
\end{align*}

For our below investigations, we will investigate the following facets: person $p$, sex $x$, session $s$, dataset collection site $d$.

# Functional

```{r, eval=FALSE}
require(fmriutils)
require(VCA)
require(stringr)

dsets = c('BNU1', 'BNU2', 'BNU3', 'HNU1', 'IBATRT', 'IPCAS1', 'IPCAS2', 'IPCAS5', 'IPCAS6', 'IPCAS8', 'NYU1', 'SWU1', 'SWU2', 'SWU3', 'SWU4', 'UWM', 'XHCUMS')

graphobj <- fmriu.io.collection.open_graphs(basepath='/data/fmri/edgelists/', datasets = dsets, atlases = c('desikan-2mm'), fmt='edgelist', rtype='array')
graphs = graphobj$graphs
```

```{r, warnings=FALSE, eval=FALSE}
gcpy = graphs
datasets = graphobj$dataset
subjects = graphobj$subjects
sessions = graphobj$sessions

parse_class <- function(basepath, dsets, subjects) {
  
  sex = list()
  disease = list()
  age = list()
  include = c()
  for (dset in dsets) {
    path_to_file = file.path(basepath, paste(dset, "_phenotypic_data.csv", sep=""))
    tab = read.csv(path_to_file)
    tab$SEX[tab$SEX == '#' | is.na(tab$SEX) | is.nan(tab$SEX)] = NaN
    tab$SEX = tab$SEX == 2
    tab = tab[complete.cases(tab$SEX),]
    tab$AGE_AT_SCAN_1 = as.numeric(levels(tab$AGE_AT_SCAN_1))[tab$AGE_AT_SCAN_1]
    for (idx in 1:dim(tab)[1]) {
      subid = toString(tab[idx,]$SUBID)
      sex[[subid]] = tab[idx,]$SEX
      age[[subid]] = tab[idx,]$AGE_AT_SCAN_1
      disease[[subid]] = tab[idx,]$DSM_IV_TR
    }
  }
  
  sclass = array(NaN, dim=c(length(subjects)))
  ageclass = sclass
  diseaseclass = sclass
  for (i in 1:length(subjects)) {
    subject = subjects[i]
    subid = sub('^0+(?=[1-9])', '', str_extract(subject, '(?<=sub-).*'), perl=TRUE)
    idx = which(names(sex) == subid)
    if (length(idx) >= 1) {
      sclass[i] <- sex[[subid]]
      ageclass[i] <- age[[subid]]
      diseaseclass[i] <- disease[[subid]]
    }
  }
  return(list(sex=sclass, age=ageclass, disease=diseaseclass))
}

sexpath = '/home/eric/Documents/research/ndmg-repos/ndmg-paper/code/fngs_class/'
class = parse_class(sexpath, dsets, subjects)
sexs = class$sex
diseases = class$disease
ages = class$age

graphs = graphs[!is.nan(ages) & !is.nan(sexs),,]
datasets = datasets[!is.nan(ages) & !is.nan(sexs)]
subjects = subjects[!is.nan(ages) & !is.nan(sexs)]
sessions = sessions[!is.nan(ages) & !is.nan(sexs)]
ager = ages[!is.nan(ages) & !is.nan(sexs)]
sexr = sexs[!is.nan(ages) & !is.nan(sexs)]
```

```{r, eval=FALSE}
VC = array(NaN, dim=c(70,70, 5))
dcoef = array(0, dim=c(70, 70))
test_dat = data.frame(sex=sexs, dataset=datasets, session=sessions, subject=subjects, age=ages, dat=array(NaN, dim=c(length(subjects))))
model = "dat ~ (1| subject) + (1| dataset) + (1| sex) + (1| session)"
var = c("subject", "dataset", "session", "sex", "residual")
for (i in 1:69) {
  print(i)
  for (j in (i+1):70) {
    print(j)
    test_dat$dat <- graphs[,i,j]
    result <- gstudy(test_dat, formula=model)
    VC[i, j,] <- result$components$var
    VC[j, i,] <- VC[i, j,]
    dcoef[i, j] <- dstudy(result, colname.objects="subject", data=test_dat, colname.scores="dat")$dependability
    dcoef[j, i] <- dcoef[i, j]
  }
}
```

Per effect, we can plot the percent variance per-edge:

```{r}
correction <- 1/apply(VC, c(1,2), sum)
for (i in 1:5){
  dat <- VC[,,i]*correction
  print(fmriu.plot.plot_graph(dat,  legend.name='percent', title=sprintf("%s variance percentage, mean=%.3f", var[i], mean(dat[!is.nan(dat)]))))
}
print(fmriu.plot.plot_graph(dcoef,  legend.name='percent', title=sprintf("dependability of model, mean=%.3f",mean(dcoef[!is.nan(dcoef)]))))
```
