---
title: "G Theory for fMRI Batch Investigations"
author: "Eric Bridgeford"
date: "October 12, 2017"
output: html_document
---


# Setting

## Task

Given:

+ $n$ samples of data-tuples, $\left\{(g_i, y_i, z_i)\right\}_{i=1}^n$.
+ A graph, $g_i \in G$, where $g_i  = (E, V, w)$ for $N=|V|$ regions of interest and $w(v_i, v_j) = w_{ij}$.
+ A set of fixed-effects, $y_i \in Y$, where $Y$ is the set of variables we expect will have an impact on connectome-to-connectome variation comprised of signal.
+ A set of random-effects, $z_i \in Z$, where $Z$ is the set of variables we expect will have an impact on connectome-to-connectome variation comprised of batch-effects.

Characterize the composition of signal to batch-effect present in the given connectomes, assuming both the fixed-effects and random-effects can be characterized as random inputs in a standard mixed-effects model.

## Statistical Model

Assume that the desired signal at a given edge can be modelled as a linear combination of the fixed and random effects under the mixed-effects model:

\begin{align*}
  w_{ij, \cdot} = Y\beta_{ij} + Zu_{ij} + \epsilon_{ij}
\end{align*}

where $w_{ij, \cdot}$ is the vector of connectome weights for the $ij^{th}$ edge. 

Then: 
+ $\mathbb{E}\left[w_{ij, \cdot}\right] = Y\beta_{ij}$
+ $\beta_{ij}$ is an unknown vector of fixed-effects
+ $u_{ij}$ is an unknown vector of random-effects where $\mathbb{E}\left[u_{ij}\right] = 0$ and $var(u) = G_{ij}$
+ $\epsilon_{ij}$ is an unknown vector of random errors, where $\mathbb{E}\left[\epsilon_{ij}\right] = 0$ and $var\left(\epsilon_{ij}\right) = R_{ij}$
+ $Y$ and $Z$ are the known design matrices relating the observations $w_{ij, \cdot}$ to $\beta_{ij}$ and $u_{ij}$

## Statistical Goal

Given that the joint density of $y$ and $u$ can be written $f(w_{ij, \cdot}, u_{ij}) = f(w_{ij, \cdot} | u_{ij}) f(u_{ij})$, assuming the normality and independence of our random-effects and errors, we can say that $u_{ij} \sim \mathcal{N}(0, G_{ij})$ and $\epsilon_{ij} \sim \mathcal{N}(0, R_{ij})$, and that $Cov(u_{ij}, \epsilon_{ij}) = 0$. Then we can use Henderson's mixed model equations:

\begin{align*}
  \begin{bmatrix}
    Y^TR^{-1}_{ij}Y & Y^TR^{-1}_{ij}Z \\
    Z^TR^{-1}_{ij}Y & Z^TR^{-1}_{ij}Z + G^{-1}_{ij} \\
  \end{bmatrix} \begin{bmatrix}
    \hat{\beta}_{ij} \\ \hat{u}_{ij}
  \end{bmatrix} &=
  \begin{bmatrix}
    Y^TR^{-1}_{ij}w_{ij, \cdot} \\
    Z^TR^{-1}_{ij}w_{ij, \cdot}
  \end{bmatrix}
\end{align*}

where $\hat{u}$ and $\hat{\beta}$ are the best linaer unbiased estimates for $\beta$ and $u$. We can solve for $\hat{\beta}$ and $\hat{u}$ using EM.

### Confidence that a given mixed-effect is significant

We are interested in the quantity of whether a given mixed-effect combinations is significant. To accomplish this, we can formulate the hypothesis of whether $H_0 := \mu_1 = ... = \mu_K$; that is, the mean activity at each edge $\mu_i$ are the same across the $i$ groupings of our given mixed-effect of interest, with $H_A := \textrm{the means are not equal}$. This can be examined using the F-test statistic:

\begin{align*}
  F &= \frac{\textrm{between-group variability}}{\textrm{within-group varaibility}} \\
  &= \frac{\sum_{i=1}^K n_i (\bar{Y}_i - \bar{Y})^2/(K - 1)}{\sum_{i=1}^K\sum_{j=1}^{n_i} (Y_{ij} - \bar{Y_{i})^2}/(N - K)}
\end{align*}

We can the compute the $p-$value where:
\begin{align*}
  p = \mathbb{P}\left\{\textrm{Falsely reject $H_0$ for $H_A \; | \; H_0$ is true}\right\}
\end{align*}

### Proportion of Total Variability accounted by a given mixed-effect

We are interested in the proportion of the variance accounted for by each variance component:

\begin{align*}
  V &= \frac{\sigma_i^2}{\sum_{j \in M}\sigma_j^2}
\end{align*}

where $M$ is the set of possible facets and between-facet interactions.

### D-Coefficient

For the univariate case, we can estimate the dependability $D$-coefficients:

\begin{align*}
  \phi_{UV}(X) = \frac{\sigma^2_p(x)}{\sigma_p(x)^2 + \sum_{i\in M \ p} \frac{\sigma_i^2}{n_i}}
\end{align*}

where $n_i$ are the number of conditions facet $i$ can occupy, and $p$ are our test-subjects in our investigation.

For our below investigations, we will investigate the following facets: person $p$, sex $x$, session $s$, dataset collection site $d$.

# Diffusion

```{r}
require(fmriutils)
require(VCA)
require(stringr)
require(gtheory)

dsets = c('BNU1', 'BNU3', 'HNU1', 'SWU4')

graphobj <- fmriu.io.collection.open_graphs(basepath='/data/dwi/edgelists/', datasets = dsets, atlases = c('desikan'), gname = "graphs", fmt='edgelist', rtype='array')
graphs = graphobj$graphs
```

```{r, warnings=FALSE}
gcpy = graphs
datasets = graphobj$dataset
subjects = graphobj$subjects
sessions = graphobj$sessions

parse_class <- function(basepath, dsets, subjects) {
  
  sex = list()
  disease = list()
  age = list()
  include = c()
  for (dset in dsets) {
    path_to_file = file.path(basepath, paste(dset, "_phenotypic_data.csv", sep=""))
    tab = read.csv(path_to_file)
    tab$SEX[tab$SEX == '#' | is.na(tab$SEX) | is.nan(tab$SEX)] = NaN
    tab$SEX = tab$SEX == 2
    tab = tab[complete.cases(tab$SEX),]
    tab$AGE_AT_SCAN_1 = as.numeric(levels(tab$AGE_AT_SCAN_1))[tab$AGE_AT_SCAN_1]
    for (idx in 1:dim(tab)[1]) {
      subid = toString(tab[idx,]$SUBID)
      sex[[subid]] = tab[idx,]$SEX
      age[[subid]] = tab[idx,]$AGE_AT_SCAN_1
      disease[[subid]] = tab[idx,]$DSM_IV_TR
    }
  }
  
  sclass = array(NaN, dim=c(length(subjects)))
  ageclass = sclass
  diseaseclass = sclass
  for (i in 1:length(subjects)) {
    subject = subjects[i]
    subid = sub('^0+(?=[1-9])', '', str_extract(subject, '(?<=sub-).*'), perl=TRUE)
    idx = which(names(sex) == subid)
    if (length(idx) >= 1) {
      sclass[i] <- sex[[subid]]
      ageclass[i] <- age[[subid]]
      diseaseclass[i] <- disease[[subid]]
    }
  }
  return(list(sex=sclass, age=ageclass, disease=diseaseclass))
}

sexpath = '/home/eric/Documents/research/ndmg-repos/ndmg-paper/code/fngs_class/'
class = parse_class(sexpath, dsets, subjects)
sexs = class$sex
diseases = class$disease
ages = class$age

graphs = graphs[!is.nan(ages) & !is.nan(sexs),,]
datasets = datasets[!is.nan(ages) & !is.nan(sexs)]
subjects = subjects[!is.nan(ages) & !is.nan(sexs)]
sessions = sessions[!is.nan(ages) & !is.nan(sexs)]
ager = ages[!is.nan(ages) & !is.nan(sexs)]
sexr = sexs[!is.nan(ages) & !is.nan(sexs)]
```

```{r}
VC = array(0, dim=c(70,70, 5))
dcoef = array(0, dim=c(70, 70))
test_dat = data.frame(sex=sexs, dataset=datasets, session=sessions, subject=subjects, age=ages, dat=array(NaN, dim=c(length(subjects))))
model = "dat ~ (1| subject) + (1| dataset) + (1| sex) + (1| session)"
var = c("subject", "dataset", "session", "sex", "residual")
for (i in 1:69) {
  for (j in (i+1):70) {
    test_dat$dat <- graphs[,i,j]
    tryCatch({
      result <- gstudy(test_dat, formula=model)
      VC[i, j,] <- result$components$var
      VC[j, i,] <- VC[i, j,]
      dcoef[i, j] <- dstudy(result, colname.objects="subject", data=test_dat, colname.scores="dat")$dependability
      dcoef[j, i] <- dcoef[i, j]},
    error=function(e) {
      VC[i,j,] <- 0
      VC[j,i,] <- 0
      dcoef[i,j] <- 0
      dcoef[j,i] <- 0
    })
  }
}
```

Per effect, we can plot the percent variance per-edge:

```{r}
dcoef[is.nan(dcoef)] <- 0
correction <- 1/apply(VC, c(1,2), sum)
correction[is.infinite(correction)] = 1
for (i in 1:5){
  dat <- VC[,,i]*correction
  print(fmriu.plot.plot_graph(dat,  legend.name='percent', title=sprintf("DWI %s variance percentage, mean=%.3f", var[i],
                                                                         mean(dat[!is.nan(dat)])), limits = c(min(dat), max(dat))))
}
print(fmriu.plot.plot_graph(dcoef,  legend.name='percent', title=sprintf("DWI dependability of model, mean=%.3f",mean(dcoef[!is.nan(dcoef)])),
                            limits=c(min(dcoef), max(dcoef))))
```

Below, we sum across dependability, obtaining the per-vertex summed dependability scores. We plot the per-vertex dependability scores, with Blue for low dependability scores and Green for higher dependability scores:

```{r}
require(RColorBrewer)
require(oro.nifti)

dvertices = apply(dcoef, 1, sum)
nvertices = apply(VC[,,2]*correction, 1, sum)

# atlas is a oro.nifti image
  plot_brain_overlay <- function(atlas, reference, values) {
    ov <- atlas
    # sort the rois so that they go low to small, which is what pipeline outputs
    roimap <- sort(unique(array(atlas@.Data)))
    roimap <- roimap[2:length(roimap)]  # ignore 0 valued roi (no region)
    for (roi in 1:length(values)) {
      if (is.nan(values[roi])) {
        ov@.Data[round(atlas@.Data) == roimap[roi]] = 0
      } else {
        ov@.Data[round(atlas@.Data) == roimap[roi]] = values[roi]
      }
    }
    overlay(reference, ifelse(ov > 0, ov, NA), zlim.y=c(min(values), max(values)),
            col.y=brewer.pal(n = 8, name = "Reds"))
  }

parc = readNIfTI('/home/eric/atlases/func_atlases/label/desikan_res-2x2x2.nii.gz')
template = readNIfTI('/home/eric/atlases/func_atlases/atlas/MNI152NLin6_t1w_res-2x2x2.nii.gz')
plot_brain_overlay(parc, template, dvertices)
```

As we can see, it appears as though the more dependable edges are the edges in the upper cortical regions. Below, we show the vertices colored by the summed (across edges similar to before) percent variance captured by the dataset covariates under the g-study:

```{r}
plot_brain_overlay(parc, template, nvertices)
```

the lower sub-cortical regions tend to show higher percent variance captured by the dataset regressors, indicating they may be more explained by batch effects than higher cortical areas.

# Functional

```{r}
require(fmriutils)
require(VCA)
require(stringr)

dsets = c('BNU1', 'BNU2', 'BNU3', 'HNU1', 'IBATRT', 'IPCAS1', 'IPCAS2', 'IPCAS5', 'IPCAS6', 'IPCAS8', 'NYU1', 'SWU1', 'SWU2', 'SWU3', 'SWU4', 'UWM', 'XHCUMS')

graphobj <- fmriu.io.collection.open_graphs(basepath='/data/fmri/raw/edgelists/', datasets = dsets, atlases = c('desikan-2mm'), fmt='edgelist', rtype='array')
graphs = graphobj$graphs
```

```{r, warnings=FALSE}
gcpy = graphs
datasets = graphobj$dataset
subjects = graphobj$subjects
sessions = graphobj$sessions

parse_class <- function(basepath, dsets, subjects) {
  
  sex = list()
  disease = list()
  age = list()
  include = c()
  for (dset in dsets) {
    path_to_file = file.path(basepath, paste(dset, "_phenotypic_data.csv", sep=""))
    tab = read.csv(path_to_file)
    tab$SEX[tab$SEX == '#' | is.na(tab$SEX) | is.nan(tab$SEX)] = NaN
    tab$SEX = tab$SEX == 2
    tab = tab[complete.cases(tab$SEX),]
    tab$AGE_AT_SCAN_1 = as.numeric(levels(tab$AGE_AT_SCAN_1))[tab$AGE_AT_SCAN_1]
    for (idx in 1:dim(tab)[1]) {
      subid = toString(tab[idx,]$SUBID)
      sex[[subid]] = tab[idx,]$SEX
      age[[subid]] = tab[idx,]$AGE_AT_SCAN_1
      disease[[subid]] = tab[idx,]$DSM_IV_TR
    }
  }
  
  sclass = array(NaN, dim=c(length(subjects)))
  ageclass = sclass
  diseaseclass = sclass
  for (i in 1:length(subjects)) {
    subject = subjects[i]
    subid = sub('^0+(?=[1-9])', '', str_extract(subject, '(?<=sub-).*'), perl=TRUE)
    idx = which(names(sex) == subid)
    if (length(idx) >= 1) {
      sclass[i] <- sex[[subid]]
      ageclass[i] <- age[[subid]]
      diseaseclass[i] <- disease[[subid]]
    }
  }
  return(list(sex=sclass, age=ageclass, disease=diseaseclass))
}

sexpath = '/home/eric/Documents/research/ndmg-repos/ndmg-paper/code/fngs_class/'
class = parse_class(sexpath, dsets, subjects)
sexs = class$sex
diseases = class$disease
ages = class$age

graphs = graphs[!is.nan(ages) & !is.nan(sexs),,]
datasets = datasets[!is.nan(ages) & !is.nan(sexs)]
subjects = subjects[!is.nan(ages) & !is.nan(sexs)]
sessions = sessions[!is.nan(ages) & !is.nan(sexs)]
ager = ages[!is.nan(ages) & !is.nan(sexs)]
sexr = sexs[!is.nan(ages) & !is.nan(sexs)]
```

```{r}
VC = array(0, dim=c(70,70, 5))
dcoef = array(0, dim=c(70, 70))
test_dat = data.frame(sex=sexs, dataset=datasets, session=sessions, subject=subjects, age=ages, dat=array(NaN, dim=c(length(subjects))))
model = "dat ~ (1| subject) + (1| dataset) + (1| sex) + (1| session)"
var = c("subject", "dataset", "session", "sex", "residual")
for (i in 1:69) {
  for (j in (i+1):70) {
    test_dat$dat <- graphs[,i,j]
    result <- gstudy(test_dat, formula=model)
    VC[i, j,] <- result$components$var
    VC[j, i,] <- VC[i, j,]
    dcoef[i, j] <- dstudy(result, colname.objects="subject", data=test_dat, colname.scores="dat")$dependability
    dcoef[j, i] <- dcoef[i, j]
  }
}
```

Per effect, we can plot the percent variance per-edge:

```{r}
correction <- 1/apply(VC, c(1,2), sum)
correction[is.infinite(correction)] = 1
for (i in 1:5){
  dat <- VC[,,i]*correction
  print(fmriu.plot.plot_graph(dat,  legend.name='percent', title=sprintf("fMRI %s variance percentage, mean=%.3f", var[i],
                                                                         mean(dat[!is.nan(dat)])), limits = c(min(dat), max(dat))))
}
print(fmriu.plot.plot_graph(dcoef,  legend.name='percent', title=sprintf("fMRI dependability of model, mean=%.3f",mean(dcoef[!is.nan(dcoef)])),
                            limits=c(min(dcoef), max(dcoef))))
```

Below, we sum across dependability, obtaining the per-vertex summed dependability scores. We plot the per-vertex dependability scores, with Blue for low dependability scores and Green for higher dependability scores:

```{r}
require(RColorBrewer)
require(oro.nifti)

dvertices = apply(dcoef, 1, sum)
nvertices = apply(VC[,,2]*correction, 1, sum)

# atlas is a oro.nifti image
  plot_brain_overlay <- function(atlas, reference, values) {
    ov <- atlas
    # sort the rois so that they go low to small, which is what pipeline outputs
    roimap <- sort(unique(array(atlas@.Data)))
    roimap <- roimap[2:length(roimap)]  # ignore 0 valued roi (no region)
    for (roi in 1:length(values)) {
      if (is.nan(values[roi])) {
        ov@.Data[round(atlas@.Data) == roimap[roi]] = 0
      } else {
        ov@.Data[round(atlas@.Data) == roimap[roi]] = values[roi]
      }
    }
    overlay(reference, ifelse(ov > 0, ov, NA), zlim.y=c(min(values), max(values)),
            col.y=brewer.pal(n = 8, name = "Reds"))
  }

parc = readNIfTI('/home/eric/atlases/func_atlases/label/desikan_res-2x2x2.nii.gz')
template = readNIfTI('/home/eric/atlases/func_atlases/atlas/MNI152NLin6_t1w_res-2x2x2.nii.gz')
plot_brain_overlay(parc, template, dvertices)
```
As we can see, it appears as though the more dependable edges are the edges in the upper cortical regions. Below, we show the vertices colored by the summed (across edges similar to before) percent variance captured by the dataset covariates under the g-study:
```{r}
plot_brain_overlay(parc, template, nvertices)
```

the lower sub-cortical regions tend to show higher percent variance captured by the dataset regressors, indicating they may be more explained by batch effects than higher cortical areas. 